\bf Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-02-10 main.Rmd
cleaning cricket and cyclic calendar algrabra

The way cyclic granularities relate become important when we consider the data structure in \@autoref{fig:data-structure}. While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others. However, irrespective of the type of cyclic granularities, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a  non-temporal dataset similar to any temporal dataset.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. The hierarchy table would look like the following:

@@

There are many interesting questions that can possibly be answered with such a data set, however, we will explore a few and understand how the proposed approach in the paper can help answer some of the questions.

@@

 cricket_per_over <- cricket_season %>%
   group_by(season,
            match_id,
            batting_team,
            bowling_team,
            inning,
            over) %>%
   summarise(runs_per_over = sum(total_runs),
             run_rate = sum(total_runs)/length(total_runs))

 cricket_tsibble_all <- cricket_per_over %>%
   ungroup() %>%
   mutate(data_index = row_number()) %>%
   as_tsibble(index = data_index)
```

@@

# cricket_tsibble %>%
#   filter(batting_team %in% c("Mumbai Indians",
#                              "Chennai Super Kings"))%>%
#   create_gran("inning_match", hierarchy_model) %>% create_gran("over_inning", hierarchy_model)

cricket_tsibble %>%
  filter(batting_team %in% c("Mumbai Indians",
                             "Chennai Super Kings"))%>%
  gran_advice("inning", 
            "over")

@@

# cricket_tsibble %>%
#   filter(batting_team %in% c("Mumbai Indians",
#                              "Chennai Super Kings"))%>%
#   gran_advice("inning", 
#             "over")

@@

@@

cricket_tsibble %>% gran_obs("over", "fielding_wckts")

cricket_data_ts <- cricket_tsibble %>% 
  mutate(field = if_else(fielding_wckts==0, "no wickets", ">0 wickets"),
         dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls"))

  #filter(fielding_wckts %in% c(0,1)) %>% 
  cricket_data_ts %>% prob_plot( "over","field",
                               hierarchy_model,
                               response = "diff_run_rate",
                               plot_type = "violin",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("") + geom_boxplot(width = 0.5)
```

@@

cricket_data_ts %>% 
  #filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("dot",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
            symmetric = FALSE)  + ggtitle("")
```

>>>>

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others.

<!-- However, irrespective of the type of cyclic granularities, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a  non-temporal dataset similar to any temporal dataset. -->

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. This could be useful in a periodic world when we wish to compute any circular/quasi-circular granularity based on a hierarchy table which look like the following:

@@

However, even if the situation is not periodic and a similar hierarchy can not be formed, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a  non-temporal dataset similar to any temporal dataset. There are many interesting questions that can possibly be answered with such a data set irrespective of the type of cyclic granularities. We will explore a few and understand how the proposed approach in the paper can help answer some of the questions.

@@

@@

@@

@@

<!-- fetching raw data since dot and fielding information not available in dismissal type not available in cricket data gravitas -->

 # cricket_per_over <- cricket_season %>%
 #   group_by(season,
 #            match_id,
 #            batting_team,
 #            bowling_team,
 #            inning,
 #            over) %>%
 #   summarise(runs_per_over = sum(total_runs),
 #             run_rate = sum(total_runs)/length(total_runs))
 # 
 # cricket_tsibble_all <- cricket_per_over %>%
 #   ungroup() %>%
 #   mutate(data_index = row_number()) %>%
 #   as_tsibble(index = data_index)
```

```{r}

@@

cricket_data<- cricket_tsibble %>% 
  mutate(field = if_else(fielding_wckts==0, "no wickets", ">0 wickets"),
         dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls")) 
```

```{r echo = TRUE}
cricket_data %>% gran_obs("field", "over")
cricket_data %>% gran_obs("dot", "over")
```

  #filter(fielding_wckts %in% c(0,1)) %>% 
  cricket_data %>% prob_plot("over","field",
                               hierarchy_model,
                               response = "diff_run_rate",
                               plot_type = "violin",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("") + geom_boxplot(width = 0.5)
```

@@

cricket_data %>% 
  #filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("dot",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
            symmetric = FALSE)  + ggtitle("")
```

### 2020-02-10 main.tex
cleaning cricket and cyclic calendar algrabra

\textbf{Hierarchy table} Let \(H_n: (G, C, P)\) be a hierarchy model containing n linear granularities. \(G_{l}\) represents the linear granularity of order l and \(P(l,m)\) represents the period length of \(G_{l}\) with respect to \(G_{m}\) and \(C_{G(l),G(m)}\) represents the cyclic granularity that relates linear granularity of order l and m, \(\forall l,m \in n, l<m\).

@@

If \(C_{G_{l},G_{m'}}(z)\) or \$C\_\{G\_\{l\},G\_\{m'\}\}(z) are multiple order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}.

@@

The way cyclic granularities relate become important when we consider the data structure in @autoref\{fig:data-structure\}. While considering data sets of the form \(<C_1, C_2, v>\), let \(S_{ij}\) be the set of index set such that for all \(s \in S_{ij}\), \(C_1(s) = A_i\) and \(C_2(s) = B_j\). Data subsets like \textless{}\(A_i\), \(B_j\), \(v(s)\)\textgreater{} can be obtained for all \(i \in {1, 2, \dots, n}\) and \(j \in {1, 2, \dots, m}\) which will lead to \(nm\) data subsets. We will discuss few cases, where one or more of these \(nm\) sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others. However, irrespective of the type of cyclic granularities, it can be interesting to visualize the distribution of a measured variable across these cyclic granularities to throw light on the periodic behavior of a non-temporal dataset similar to any temporal dataset.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from \href{https://www.kaggle.com/josephgpinto/ipl-data-analysis/data}{Kaggle}. The \texttt{cricket} data set in the \texttt{gravitas} package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. The hierarchy table would look like the following:

@@

\begin{verbatim}
#> The chosen granularities are harmonies 
#>  
#> Recommended plots are: violin lv quantile boxplot 
#>  
#> Number of observations are homogenous across facets 
#>  
#> Number of observations are homogenous within facets 
#>  
#> Cross tabulation of granularities : 
#>  
#> # A tibble: 20 x 3
#>    over    `1`   `2`
#>    <fct> <dbl> <dbl>
#>  1 1        65    44
#>  2 2        65    44
#>  3 3        65    44
#>  4 4        65    44
#>  5 5        65    44
#>  6 6        65    44
#>  7 7        65    44
#>  8 8        65    44
#>  9 9        65    44
#> 10 10       65    44
#> 11 11       65    44
#> 12 12       65    44
#> 13 13       65    44
#> 14 14       65    44
#> 15 15       65    44
#> 16 16       65    44
#> 17 17       65    44
#> 18 18       65    44
#> 19 19       65    44
#> 20 20       65    44
\end{verbatim}

@@

\begin{verbatim}
#> # A tibble: 4 x 21
#>   fielding_wckts   `1`   `2`   `3`   `4`   `5`   `6`   `7`   `8`   `9`  `10`
#>   <fct>          <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 0               1039   997   972   978   962   957  1002  1012   993   984
#> 2 1                120   145   171   165   180   181   138   124   140   153
#> 3 2                  5    10     8     8     9    10     6    10     8     4
#> 4 3                  0     0     0     0     0     0     0     0     0     0
#> # … with 10 more variables: `11` <dbl>, `12` <dbl>, `13` <dbl>, `14` <dbl>,
#> #   `15` <dbl>, `16` <dbl>, `17` <dbl>, `18` <dbl>, `19` <dbl>, `20` <dbl>
\end{verbatim}

>>>>

\textbf{Hierarchy table} Let \(H_n: (G, C, P)\) be a hierarchy model containing n linear granularities. \(G_{l}\) represents the linear granularity of order l and \(P(l,m)\) represents the period length of \(G_{l}\) with respect to \(G_{m}\) and \(C_{G(l),G(m)}\) represents the cyclic granularity that relates linear granularity of order l and m, \(\forall l,m \in {1, 2, \dots,n}\) and \(l<m\).

@@

If \(C_{G_{l},G_{m'}}(z)\) or \(C_{G_{l},G_{m'}}(z)\) are multiple order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \ref{fig:data-structure}. While considering data sets of the form \(<C_1, C_2, v>\), let \(S_{ij}\) be the set of index set such that for all \(s \in S_{ij}\), \(C_1(s) = A_i\) and \(C_2(s) = B_j\). Data subsets like \textless{}\(A_i\), \(B_j\), \(v(s)\)\textgreater{} can be obtained for all \(i \in {1, 2, \dots, n}\) and \(j \in {1, 2, \dots, m}\) which will lead to \(nm\) data subsets. We will discuss few cases, where one or more of these \(nm\) sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

The method is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized to other applications. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. An idea of cyclic granularities can be conceived with this hierarchy in mind. Examples may include ball of the over, over of the innings or ball of the innings. Although most of these cyclic granularities are circular in theory by design of the hierarchy, in reality these may be aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-the-over will be circular in most cases and aperiodic in others.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from \href{https://www.kaggle.com/josephgpinto/ipl-data-analysis/data}{Kaggle}. The \texttt{cricket} data set in the \texttt{gravitas} package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each inning has 20 overs and each match has 2 innings. This could be useful in a periodic world when we wish to compute any circular/quasi-circular granularity based on a hierarchy table which look like the following:

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cricket_data }\OperatorTok{%>%}\StringTok{ }\KeywordTok{gran_obs}\NormalTok{(}\StringTok{"field"}\NormalTok{, }\StringTok{"over"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> # A tibble: 20 x 3
#>    over  `>0 wickets` `no wickets`
#>    <fct>        <dbl>        <dbl>
#>  1 1              125         1039
#>  2 2              155          997
#>  3 3              179          972
#>  4 4              173          978
#>  5 5              189          962
#>  6 6              191          957
#>  7 7              144         1002
#>  8 8              134         1012
#>  9 9              148          993
#> 10 10             157          984
#> 11 11             173          962
#> 12 12             177          951
#> 13 13             169          954
#> 14 14             205          913
#> 15 15             208          892
#> 16 16             219          865
#> 17 17             231          833
#> 18 18             291          743
#> 19 19             287          695
#> 20 20             306          572
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cricket_data }\OperatorTok{%>%}\StringTok{ }\KeywordTok{gran_obs}\NormalTok{(}\StringTok{"dot"}\NormalTok{, }\StringTok{"over"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> # A tibble: 20 x 3
#>    over  `>0 dot balls` `no dot balls`
#>    <fct>          <dbl>          <dbl>
#>  1 1               1146             18
#>  2 2               1140             12
#>  3 3               1111             40
#>  4 4               1112             39
#>  5 5               1099             52
#>  6 6               1099             49
#>  7 7               1057             89
#>  8 8               1026            120
#>  9 9               1027            114
#> 10 10              1010            131
#> 11 11               992            143
#> 12 12               971            157
#> 13 13               970            153
#> 14 14               957            161
#> 15 15               935            165
#> 16 16               904            180
#> 17 17               881            183
#> 18 18               852            182
#> 19 19               788            194
#> 20 20               683            195
\end{verbatim}
